---
format:
  html:
    author: "OndÅ™ej Mottl"
    toc: true
    keep-md: true
    code-link: true
    embed-resources: true
    code-line-numbers: true
    theme: [default, custom_theme_exercise.scss]
---

# Basic functions of neotoma2 : Working with pollen data

Simple example of woking with pollen data using the [{neotoma2} package](https://open.neotomadb.org/neotoma2/).

## Setup

```{r}
#| label: setup
#| include: true
#| output: false

# load libraries
library(tidyverse) # general data wrangling and visualisation âœ¨
library(neotoma2) # # access to the Neotoma database ðŸŒ¿
library(pander) # nice tables ðŸ˜
library(here) # for working directory ðŸ—ºï¸
library(janitor) # string cleaning ðŸ§¹
library(geojsonsf) # geojson spatial data ðŸŒ
library(here) # for working directory ðŸ—ºï¸

# set the working directory
here::i_am("R/Exercises/01_neotoma2_basics.qmd")

# source the config file
source(
  here::here("R/00_Confiq_file.R")
)

# quarto render options
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 7,
  fig.align = "center",
  out.width = "100%",
  echo = TRUE
)
```

## Sites

```{r}
#| label: 2.0.
#| eval: false
#| echo: true
# check the documentation
?neotoma2::get_sites()
```

### Search by IDs

Search for site by a single numeric ID:
```{r}
#| label: 2.1.1
neotoma2::get_sites(15799) %>% 
  plot_table()
```

Search for sites with multiple IDs:
```{r}
#| label: 2.1.2
neotoma2::get_sites(
  c(15799, 15683)
) %>% 
  plot_table()
```

Searching for Sites by Name. Notet that  `%` is a wildcard character:

```{r}
#| label: 2.1.3
neotoma2::get_sites(sitename = "Alex%") %>% 
  plot_table()
```

### Searching for Sites by Age

Record span at least 8200 years:

```{r}
#| label: 2.2.1
neotoma2::get_sites(
  ageof = 8200,
  all_data = FALSE # this will only show 25 records
) %>% 
 plot_table(head = TRUE)
```

Record must PARTLY span the age range

```{r}
#| label: 2.2.2
neotoma2::get_sites(
  minage = 5000,
  maxage = 8000,
  all_data = FALSE # this will only show 
) %>% 
 plot_table(head = TRUE)
```

Record must COMPLETELY span the age range

```{r}
#| label: 2.2.3
neotoma2::get_sites(
  ageyounger = 5000,
  ageolder = 8000,
  all_data = FALSE # this will only show 25 records
) %>% 
 plot_table(head = TRUE)
```

## Datasets

You can search by all the same age properties as for sites (`ageof`, `minage`, `maxage`, `ageyounger`, `ageolder`).

```{r}
#| label: 3.0.
#| eval: false
#| echo: true
# check the documentation
?neotoma2::get_datasets()
```

### Search by IDs

```{r}
#| label: 3.1.
neotoma2::get_datasets(
  c(5, 10, 15, 20)
) %>% 
 plot_table(head = TRUE)
```

### Search by type
```{r}
#| label: 3.2.
neotoma2::get_datasets(
  all_data = FALSE, # running with `all_data = TRUE` is heavy on the server
  datasettype = "pollen"
) %>% 
 plot_table(head = TRUE)
```

### Search by geo location

Go to [geojson.io](https://geojson.io/) and get the coordinates of a polygon.

For example:
```{r}
#| label: JSON example
#| eval: false
{
        "type": "FeatureCollection",
        "features": [
          {
            "type": "Feature",
            "properties": {},
            "geometry": {
              "coordinates": [
                [
                  ...
                ]
              ],
              "type": "Polygon"
            }
          }
        ]
      }
```

Now, we can use the coordinates to search for datasets:

```{r}
#| label: 3.3.1
sel_polygon <-
  geojsonsf::geojson_sf(
    '{
        "type": "FeatureCollection",
        "features": [
          {
            "type": "Feature",
            "properties": {},
            "geometry": {
              "coordinates": [
                [
                  [
                    8.885566596626347,
                    49.771446037909755
                  ],
                  [
                    10.222591262669596,
                    48.31793402541106
                  ],
                  [
                    12.353498585013057,
                    47.56711633708565
                  ],
                  [
                    16.327096438141012,
                    48.350003278892444
                  ],
                  [
                    16.435889807717132,
                    49.83389459390426
                  ],
                  [
                    16.022485237853516,
                    51.17135995959822
                  ],
                  [
                    14.214447964984942,
                    51.55099235084026
                  ],
                  [
                    10.091782211463311,
                    50.45895941972614
                  ],
                  [
                    9.223835080442456,
                    50.26942261795418
                  ],
                  [
                    8.885566596626347,
                    49.771446037909755
                  ]
                ]
              ],
              "type": "Polygon"
            }
          }
        ]
      }'
  )

neotoma2::get_datasets(
  loc = sel_polygon
) %>% 
 plot_table()
```

### Filter

You can additionaly filter the compilation based on `lat`, `long`, `altitude`, `age_range_young`, and/or `age_range_old`

```{r}
#| label: 3.4.0
#| eval: false
#| echo: true
# check the documentation
?neotoma2::filter()
```

```{r}
#| label: 3.4.1
neotoma2::get_datasets(
  loc = sel_polygon
) %>%
  neotoma2::filter(
    datasettype == "pollen" &
      altitude > 500 &
      age_range_young <= 1e3
  ) %>% 
 plot_table()
```

## Downloading data

### Download individual record

Let's download a record with `datasetid` 24279

```{r}
#| label: 4.1.
neotoma2::get_downloads(24279) %>% 
 plot_table()
```

### Download multiple records

Download all records by sites

```{r}
#| label: 4.2.1
# get sites information
neotoma2::get_sites(sitename = "Alex%") %>%
  # get datasets information
  neotoma2::get_datasets() %>%
  # get downloads
  neotoma2::get_downloads() %>% 
  plot_table()
```

Download all records by datasets

```{r}
#| label: 4.2.2
# get datasets information
neotoma2::get_datasets(
  loc = sel_polygon
) %>%
  # filter datasets
  neotoma2::filter(
    datasettype == "pollen" &
      altitude > 500 &
      age_range_young <= 1e3
  ) %>%
  # get downloads
  neotoma2::get_downloads() %>% 
  plot_table()
```

## Working with pollen counts

### Get samples

download all records by datasets
```{r}
#| label: 5.1.1
data_selected_downloads <-
  neotoma2::get_datasets(
    loc = sel_polygon
  ) %>%
  # filter datasets
  neotoma2::filter(
    datasettype == "pollen" &
      altitude > 500 &
      age_range_young <= 1e3
  ) %>%
  # get downloads
  neotoma2::get_downloads()
```

```{r}
#| label: 5.1.0
#| eval: false
#| echo: true
# check the documentation
?neotoma2::samples()
```

Extraxt Sample information
```{r}
#| label: 5.1.2
data_selected_samples <-
  neotoma2::samples(data_selected_downloads) %>%
  as.data.frame() %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    datasetid_sampleid = paste0(datasetid, "_", sampleid)
  )

plot_table(data_selected_samples[1:5, 1:5])
```

### Get pollen counts

```{r}
#| label: 5.2.0
#| eval: false
#| echo: true
# check the documentation
?neotoma2::taxa()
```

Get vector of all "pollen" taxa
```{r}
#| label: 5.2.
vec_taxa_pollen <-
  neotoma2::taxa(data_selected_downloads) %>%
  dplyr::filter(element == "pollen") %>%
  purrr::pluck("variablename") %>%
  sort()

head(vec_taxa_pollen)
```

Get pollen counts
```{r}
#| label: 5.3.
data_sample_pollen_counts <-
  data_selected_samples %>%
  dplyr::select("datasetid_sampleid", "value", "variablename") %>%
  # only include selected taxons
  dplyr::filter(
    variablename %in% vec_taxa_pollen
  ) %>%
  dplyr::arrange(variablename) %>%
  # turn into the wider format
  tidyr::pivot_wider(
    names_from = "variablename",
    values_from = "value",
    values_fill = 0
  ) %>%
  # clean names
  janitor::clean_names()

plot_table(data_sample_pollen_counts[1:5, 1:5]) 
```

### Getting the age information 

```{r}
#| label: 6.1.
data_sample_age <-
  data_selected_samples %>%
  dplyr::distinct(datasetid_sampleid, depth, age) %>%
  dplyr::arrange(datasetid_sampleid, age)

plot_table(data_sample_age, head = TRUE)
```

### Plotting pollen diagram

Data wrangling
```{r}
#| label: 7.1.
data_to_plot <-
  data_sample_pollen_counts %>%
  # we need to turn the pollen counts into percentages
  tibble::column_to_rownames("datasetid_sampleid") %>%
  dplyr::mutate(
    colsum = rowSums(.)
  ) %>%
  dplyr::mutate(
    dplyr::across(
      -(colsum),
      ~ (. / colsum) * 100
    )
  ) %>%
  tibble::rownames_to_column("datasetid_sampleid") %>%
  dplyr::select(-colsum) %>%
  # turn into the longer format
  tidyr::pivot_longer(
    cols = -datasetid_sampleid,
    names_to = "taxon",
    values_to = "count"
  ) %>%
  # add age information
  dplyr::left_join(
    data_sample_age,
    by = "datasetid_sampleid"
  ) %>%
  dplyr::select(
    "datasetid_sampleid", "age", "taxon", "count"
  ) %>%
  tidyr::separate(
    datasetid_sampleid,
    c("datasetid", "sampleid"),
    sep = "_"
  )

plot_table(data_to_plot, head = TRUE)
```

Get the most common taxa
```{r}
#| label: 7.2.
vec_common_taxa <-
  data_to_plot %>%
  dplyr::group_by(datasetid, taxon) %>%
  dplyr::summarise(
    .groups = "drop",
    mean_count = mean(count, na.rm = TRUE)
  ) %>%
  dplyr::group_by(datasetid) %>%
  dplyr::arrange(
    dplyr::desc(mean_count)
  ) %>%
  dplyr::slice(1:10) %>%
  dplyr::ungroup() %>%
  dplyr::pull(taxon) %>%
  unique()
```

```{r}
#| label: 7.3.
data_to_plot %>%
  dplyr::filter(
    taxon %in% vec_common_taxa
  ) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = age,
      y = count,
      fill = taxon,
      col = taxon
    )
  ) +
  ggplot2::labs(
    x = "Age (cal yr BP)",
    y = "Pollen count (%)",
    title = "Pollen diagrams - Proportion of pollen taxa over time",
    subtitle = "Only 10 most common taxa per record are shown",
    caption = paste(
      "Data from Neotoma database for",
      length(unique(data_to_plot$datasetid)),
      "records"
    )
  ) +
  ggplot2::scale_x_reverse(
    breaks = scales::pretty_breaks(n = 10)
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 100),
    breaks = scales::pretty_breaks(n = 5)
  ) +
  ggplot2::theme(
    legend.position = "none",
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  ) +
  ggplot2::facet_grid(datasetid ~ taxon) +
  ggplot2::coord_flip() +
  ggplot2::geom_ribbon(
    mapping = ggplot2::aes(
      ymin = 0,
      ymax = count
    ),
    alpha = 0.5
  ) +
  ggplot2::geom_line()
```
